A Simple Minibatch Acceptance Test for MCMC

### 2016-05-19 nips_2016.tex
Did lots of work on Section 5.1

We conduct three sets of experiments to explore the benefits of our minibatch MH test and to
benchmark it with previous work. In Section~\ref{ssec:gaussians}, we show that our test enables
samples to converge to the posterior distribution of a Gaussian mixture model. In
Section~\ref{ssec:logistic}, we analyze its efficiency on logistic regression. In
Section~\ref{ssec:nets}, we apply the test for deep learning.
Appendices~\ref{app:gaussian},~\ref{app:logistic}, and~\ref{app:nnet} contain more detailed
information on these experiments.

@@

We start with a simple mixture of Gaussians, which have tied means:
\[
\theta_1 \sim N(0, \sigma_1^2); \theta_2 \sim N(0, \sigma_2^2); x_i \sim \frac{1}{2}N(\theta_1, \sigma_x^2) + \frac{1}{2} N(\theta_2, \sigma_x^2)
\]
where $\sigma_1^2 = 10, \sigma_2^2 = 1$ and $\sigma_x^2=2$. 200 data points are generated from the
model with $\theta_1 = 0$ and $\theta_2 = 1$. The theoretical posterior distribution can be seen in
Fig.~3a. We use random walk as the proposer. We set the step size 0.1 as a constant. We compare
three scenarios: with our mini-batch MH testing (see Fig.~3b), without our mini-batch MH testing
(see Fig.~3c), and with the standard MH testing (see Fig.~3d).

\begin{figure}[t]
    \centering
    \includegraphics[width=1\linewidth]{cloud_v01.png}
    \caption{
    {\color{blue}
    Daniel: I'll fix the caption later. Basically, true log posterior, standard/adaptive procedures,
    and our method, using MB=200 out of the 10k samples here (also 10k passes over data). All three
    used the same proposal distribution, a RW with a Gaussian of 0.03. Temperature is at 110.
    }
    }
    %\vspace{-10pt}
\end{figure}

\begin{figure}[t]
    \centering
    \includegraphics[width=1\linewidth]{adaptive_and_ours_information_v01.png}
    \caption{
    {\color{blue}
    Daniel: I'll fix the caption later. First plot is capped at 1k but the first bin goes up to
    8000. For the SDs, I skipped the ones that didn't satisfy the requirement (a.k.a. repeating the
    previous sample). The third is when I re-ran the experiments but with different minibatches, to
    see how much I would have to repeat. These values were all with one run, though, I might
    consider just taking several (about five?) and doing error bar stuff.
    }
    }
    %\vspace{-10pt}
\end{figure}

From the figures above, we find that with the help of our mini-batch MH testing we can obtain the
sampled distribution which are very close to target one. We feature draw our estimation of standard
derivation of $X_{\rm norm}$ over iterations, see Fig.~3e. We find with even constant step size, the
standard derivation decreases, because the gradient of posterior distribution declines and make the
scale of $\Delta'$ deminished when the sampled point is apporching MAP. Thus, the error our
transition kernel will declines with even constant step size as discribed in Section~\ref{sec:theory}.

@@

In this paper, we have derived a new MH test for minibatch MCMC methods. We demonstrated how a
simple deconvolution process allows us to use a minibatch approximation to the full data tests. We
experimentally show the benefits our MH test.

Straightforward directions for future work include running additional experiments, with a particular
focus on investigation of the variance precondition.  More elaborate extensions include combining
our results with Hamiltonian Monte Carlo methods, providing a recipe for how to use our algorithm
(following the general framework of~\cite{sgmcmc_2015}), or integrating parallel
MCMC~\cite{conf/uai/AngelinoKWSA14,conf/icml/AhnSW14} concepts into our work.

{\color{blue}
Daniel: TODO, clean this up!
}

>>>>

We conduct three sets of experiments to explore the benefits of our minibatch MH test and to
benchmark it with previous work. In Section~\ref{ssec:gaussians}, we show that our test enables
samples to converge to the posterior distribution of a heated Gaussian mixture model. In
Section~\ref{ssec:logistic}, we analyze its efficiency on logistic regression. In
Section~\ref{ssec:nets}, we apply the test for deep learning.
Appendices~\ref{app:gaussian},~\ref{app:logistic}, and~\ref{app:nnet} contain more detailed
information on these respective experiments.

@@

We start with a simple Gaussian mixture model, borrowing an experiment from~\cite{langevin_2011}.
The parameter is 2-D, $\theta = (\theta_1,\theta_2)$, and the parameter/data generation process is
\begin{equation}\label{eq:data_generation}
(\theta_1,\theta_2) \sim \mathcal{N}((0,0), {\rm diag}(\sigma_1^2, \sigma_2^2));\quad \quad x_i \sim
0.5 \cdot \mathcal{N}(\theta_1, \sigma_x^2) + 0.5 \cdot \mathcal{N}(\theta_1+\theta_2, \sigma_x^2).
\end{equation}
We set $\sigma_1^2 = 10, \sigma_2^2 = 1$ and $\sigma_x^2=2$. Fixing $\theta = (0,1)$, we draw 10000
data points so that the target distribution is $p(\theta)\prod_{i=1}^{10000}p(x_i\mid \theta)$, with
the prior based on the $\theta$ generation process in Equation~\ref{eq:data_generation}. This turns
out to be a rather sharp posterior, so to make our test applicable, we apply a temperature $T=110$
to reduce ${\rm std}(\Delta')$. Taking logs, we get the new target as shown in the far left of
Figure~\ref{fig:gauss_mix_1}.

We run MCMC with our MH test with $m=200$. We also run this using standard full-batch MCMC
(``standard sampling'') with the standard MH test, and the method from~\cite{cutting_mh_2014}
(``adaptive sampling''). For the latter, we also use $m=200$ and increment minibatches by that
amount within an iteration if necessary. The tolerance for making a decision is $\epsilon=0.1$. To
make comparisons easier, all three use the same random walk proposer with covariance $\Sigma={\rm
diag}(0.03, 0.03)$. This is a poor proposer, but it is sufficient for our purposes as the quality of
the samples will be due to the MH test. All methods are run 10000 times to collect 10000 samples.

\begin{figure}[t]
    \centering
    \includegraphics[width=1\linewidth]{cloud_v01.png}
    \caption{
    The log posterior contours (temperature 110) and three scatter plots of sampled $\theta$ values.
    }
    \label{fig:gauss_mix_1}
    %\vspace{-10pt}
\end{figure}

\begin{figure}[t]
    \centering
    \includegraphics[width=1\linewidth]{adaptive_and_ours_information_v01.png}
    \caption{
    Histograms of minibatch sizes and ${\rm std}(\Delta')$ values, along with ``${\rm std}(\Delta')$
    violations'' (right).
    }
    \label{fig:gauss_mix_2}
    %\vspace{-10pt}
\end{figure}

Figure~\ref{fig:gauss_mix_1} shows three scatter plots of the resulting $\theta$ samples for the
three methods, with darker regions indicating a greater density of points. All three methods obtain
the same rough form of the posterior, so our MH test can indeed obtain the posterior just as the
other methods do.

Figure~\ref{fig:gauss_mix_2} (to the left) shows a histogram of the final minibatch sizes used by
the adaptive subsampling method each iteration (capped at 600 for readability).  Usually, the method
can make a decision with 200 samples. Occasionally, however, it must use all 10000 points, resulting
in roughly 50 times more computation. For this particular experiment, that happened eight times.  In
contrast, our method keeps the minibatch size fixed at 200, but requires that ${\rm std}(\Delta') <
1.2$. Figure~\ref{fig:gauss_mix_2} (in the middle) shows a histogram of the estimated ${\rm
std}(\Delta')$ values each iteration. Only 75 iterations resulted in ${\rm std}(\Delta') \ge 1.2$
and for those, we skipped the rest of the iteration (setting $\theta_{t+1}=\theta_t$). To smooth the
${\rm std}(\Delta')$ values, we use moving average updates. The third plot in
Figure~\ref{fig:gauss_mix_2} investigates the number of times ${\rm std}(\Delta') \ge 1.2$. For six
minibatch sizes (50, 100, 150, 200, 250, and 300), we ran MCMC with our MH test five times and the
number of violations averaged. As minibatch sizes increases, the number of violations decreases,
showing that our test is more applicable.

@@

In this paper, we have derived a new MH test for minibatch MCMC methods. We demonstrated how a
simple deconvolution process allows us to use a minibatch approximation to the full data tests. We
experimentally show the benefits of our test on Gaussian mixtures, logistic regression, and deep
learning.  Straightforward directions for future work include running more experiments with a
particular focus on investigation of the variance precondition.  More elaborate extensions include
combining our results with Hamiltonian Monte Carlo methods, providing a recipe for how to use our
algorithm (following the framework of~\cite{sgmcmc_2015}), or integrating parallel
MCMC~\cite{conf/uai/AngelinoKWSA14,conf/icml/AhnSW14} concepts.
